Gravitational wave sources in our Galactic backyard: Predictions for BH and NS binaries in LISA

### 2021-07-21 paper/tex/sections/methods.tex
updates to 3.1 and 3.2 from Selma

We define a binary as detectable if its gravitational wave signal has a signal-to-noise ratio of greater than 7 \citep[e.g.][]{Breivik+2020, Korol+2020}. The sky-, polarisation- and orientation-averaged signal-to-noise ratio, $\rho$, of an inspiraling binary can be calculated with the following \citep[e.g.][]{Finn+2000}
\begin{equation}\label{eq:snr}
    \rho^2 = \sum_{n=1}^{\infty} \int_{f_{n, i}}^{f_{n, f}} \frac{h_{c, n}^{2}}{f_{n}^{2} S_{\rm n}\left(f_{n}\right)} \dd{f_n},
\end{equation}
where $n$ is a harmonic of the gravitational wave signal, $f_n = n \cdot f_{\rm orb}$ is the frequency of the $n^{\rm th}$ harmonic of the gravitational wave signal, $f_{\rm orb}$ is the orbital frequency, $S_{\rm n}(f_n)$ is the LISA sensitivity curve at frequency $f_n$ \citep[e.g.][]{Robson+2019} and $h_{c,n}$ is the characteristic strain of the $n^{\rm th}$ harmonic, given by \citep[e.g.][]{Barack+2004}
\begin{equation}\label{eq:charstrain}
    h^2_{c,n} = \frac{2^{5/3}}{3 \pi^{4/3}} \frac{(G \mathcal{M}_c)^{5/3}}{c^3 D_L^2} \frac{1}{f_{\rm orb}^{1/3}} \frac{g(n,e)}{n F(e)},
\end{equation}
where $D_L$ is the luminosity distance to the source, $f_{\rm orb}$ is the orbital frequency, $g(n, e)$ and $F(e)$ are given in \citet{Peters+1963} and $\mathcal{M}_c$ is the chirp mass, defined as
\begin{equation}\label{eq:chirp_mass}
    \mathcal{M}_c = \frac{(m_1 m_2)^{3/5}}{(m_1 + m_2)^{1/5}}.
\end{equation}

>>>>

We define a binary as detectable if its gravitational wave signal has a signal-to-noise ratio (SNR) of greater than 7 \citep[e.g.][]{Breivik+2020, Korol+2020}. The sky-, polarisation- and orientation-averaged signal-to-noise ratio, $\rho$, of an inspiraling binary can be calculated with the following \citep[e.g.][]{Finn+2000}
\begin{equation}\label{eq:snr}
    \rho^2 = \sum_{n=1}^{\infty} \int_{f_{n, i}}^{f_{n, f}} \frac{h_{c, n}^{2}}{f_{n}^{2} S_{\rm n}\left(f_{n}\right)} \dd{f_n},
\end{equation}
where $n$ is a harmonic of the gravitational wave signal, $f_n = n \cdot f_{\rm orb}$ is the frequency of the $n^{\rm th}$ harmonic of the gravitational wave signal, $f_{\rm orb}$ is the orbital frequency, $S_{\rm n}(f_n)$ is the LISA sensitivity curve at frequency $f_n$ \citep[e.g.][]{Robson+2019} and $h_{c,n}$ is the characteristic strain of the $n^{\rm th}$ harmonic, given by \citep[e.g.][]{Barack+2004}
\begin{equation}\label{eq:charstrain}
    h^2_{c,n} = \frac{2^{5/3}}{3 \pi^{4/3}} \frac{(G \mathcal{M}_c)^{5/3}}{c^3 D_L^2} \frac{1}{f_{\rm orb}^{1/3}} \frac{g(n,e)}{n F(e)},
\end{equation}
where $D_L$ is the luminosity distance to the source, $f_{\rm orb}$ is the orbital frequency, $g(n, e)$ and $F(e)$ are given in \citet{Peters+1963} and $\mathcal{M}_c$ is the chirp mass, defined as
\begin{equation}\label{eq:chirp_mass}
    \mathcal{M}_c = \frac{(m_1 m_2)^{3/5}}{(m_1 + m_2)^{1/5}}.
\end{equation}

### 2021-07-21 paper/tex/sections/results.tex
updates to 3.1 and 3.2 from Selma

In this section we present our main results for the detectable LISA DCO population in our fiducial model. We find that on average, a 4-year LISA mission will detect about \BHBHFourYear{} BHBHs, \BHNSFourYear{} BHNSs and \NSNSFourYear{} NSNSs (c.f.\ Table~\ref{tab:detection_rates}). We first show the distribution of the sources together with the sensitivity curve in Section~\ref{sec:dcos_on_sc}, before exploring the parameter distributions for detectable sources in Section~\ref{sec:fiducial_distributions} and the measurement uncertainties in Section.~\ref{sec:measurement_uncertainties}.

\subsection{Distribution on the sensitivity curve}\label{sec:dcos_on_sc}

\begin{figure*}[p]
    \centering
    \includegraphics[width=\textwidth]{2_dcos_on_sc.png}
    \caption{Density distribution of detectable BHBH, BHNS and NSNS binaries are shown together with the LISA sensitivity curve. In the top panel we show all systems with the LISA verification binaries over plotted (star symbols, \citealp{Kupfer+2018}). In the bottom panels we separate by type. Contours show the percentage of the population enclosed. The remaining 2\% of the population is shown as dots with a size that scales with the sampling weight. For reference we show lines where a circular binary of average chirp mass $\avg{\mathcal{M}_c}$ would reside for a given remaining inspiral time (vertical lines) and distance (diagonal line). To highlight the role of eccentricity we further show the signal expected for an eccentric binary at 30 kpc. The coloured line in the bottom panels shows a contour that encloses 90\% of the population that is circular. See Sec.~\ref{sec:dcos_on_sc} for a discussion.}
    \label{fig:dcos_on_sc}
\end{figure*}

We show the expected distribution of detectable DCOs together with the LISA sensitivity curve in Figure~\ref{fig:dcos_on_sc}. This shows that the detectable population of these DCOs is concentrated at comparatively lower frequencies than the LISA verification binaries (shown as stars in the top panel). This is expected since producing the same SNR as a BHBH, BHNS or NSNS with a relatively lower mass (circular) WDWD requires a higher frequency. This finding is in agreement with \citet{Sesana+2020} and as noted in that work, this could possibly be used to distinguish more massive DCOs from WDWDs probabilistically. There are several other notable features in these distributions. To understand these we overplot reference lines indicating where a circular binary with the average chirp mass ($\avg{\mathcal{M}_c}$, annotated in each panel) would reside for a given remaining inspiral time (vertical lines) and a given distance (diagonal lines).

Firstly, we note that the peak of the density distribution coincides with the centre of the Milky Way as expected, since binaries are most likely to be formed towards the centre of the Galaxy. Additionally, we expect that if a population is entirely circular, it should be bounded approximately between the $0.1$--$30 \unit{kpc}$ lines, roughly the minimum and maximum distance to a source in the Milky Way. From inspection of the bottom panels with each individual DCO type we see that this is the true for a large fraction of the population. However, there is a distinct subpopulation of binaries that extend downwards, especially around $2 \unit{mHz}$. This offshoot is composed of eccentric binaries for which the circular distance contours do not apply. For illustrative purposes, we plot the 90\% contour of only the \textit{circular} sources in our sample over the density distribution in each of the bottom panels and this subsample is indeed bounded by $0.1$--$30 \unit{kpc}$ the distance lines. We also plot a line of constant distance at $30 \unit{kpc}$ for an eccentric binary with $e = 0.98$ to show the different limit for eccentric sources. To further illustrate this point, we show the same plot but with each point coloured by its eccentricity in Fig.~\ref{fig:dcos_on_sc_ecc_col}. Overall, we can see that eccentric sources tend to be located on a slightly different location on the sensitivity curve and so this could perhaps be used help in identifying them.

We plot vertical lines that give the inspiral time for a circular binary with the average chirp mass (annotated in each panel). From these lines we can understand the trend of the density distribution decreasing with increasing frequency. Sources with higher frequencies have shorter inspiral times and thus DCOs will spend less time in these regimes, meaning that more sources are detected at lower frequencies. Note that these inspiral time lines should only be used as guidelines for the population as a whole, as the inspiral time of each individual source will be a function of its mass and eccentricity. It is also evident for each DCO source that the tail of the high frequency sources is more numerous near to the Galactic centre than at short distances. This is simply because there are more sources in the galactic centre and so the chances of `catching' a binary at high frequency are better.

\subsection{Properties of the detectable systems}\label{sec:fiducial_distributions}

In Figure~\ref{fig:fiducial_pdf_distributions}, we show the distribution of the individual parameters of the population of detectable binaries and discuss the various features in the following sections.

\subsubsection{Orbital Frequency}
The orbital frequency distributions for BHBHs, BHNSs and NSNSs peak at progressively increasing frequencies. This is because a higher mass DCO at the same distance and eccentricity requires a lower frequency to produce the same signal-to-noise ratio and thus be detected. The BHBH distribution has a tail that extends to $8 \times 10^{-6} \unit{Hz}$, which is comprised of highly eccentric binaries. These systems are still detectable by LISA as the high eccentricity means that the majority of the GW signal is emitted at higher harmonics at higher frequencies that are located in the LISA band. Similar tails are not as prevalent for BHNSs and NSNSs as they do not have as many eccentric binaries.

@@

\subsubsection{Black Hole Mass}
For both the BHBHs and BHNSs, the black hole mass distribution extends to relatively low masses, with $\BHBHmBHBelowEleven{}$ and $\BHNSmBHBelowEleven{}$ respectively below $11 \unit{M_{\odot}}$. Unlike ground-based detectors, LISA is not biased to higher masses and so the mass distribution more closely follows the IMF. In addition, at the high metallicities in the Milky Way, stellar winds are much stronger and strip away much of the stellar mass before BH formation, resulting the less massive black holes. The mass distribution extends down to $2.5 \unit{M_{\odot}}$, our fiducial maximum neutron star mass. Since the \citet{Fryer+2012} \textit{delayed} remnant mass prescription (applied in this work) does not produce a mass gap between neutron stars and black holes. Indeed we expect $\BHBHLowerMassGap{}$ and $\BHNSLowerMassGap{}$ of detected BHBH and BHNS systems to contain a black hole in the lower mass gap. Therefore, LISA could help to confirm or rule out the existence of the lower mass gap.

@@

\subsubsection{Mass Ratio}
The mass ratio distribution for each DCO type is relatively distinct from the others. The majority of NSNSs have a mass ratio close to unity, with $\NSNSqAbovePointEight{}$ of systems having $q > 0.8$. The reason for the concentration around equal masses is that most NSs are formed either through electron-capture supernovae or from low mass stars. We set the remnant mass for any system formed through ECSN to $1.26 \unit{M_{\odot}}$ (see Sec.~\ref{sec:fiducial_physics}). The remnant mass prescription that we use gives a fixed fallback mass for any star with a CO core mass less than $2.5 \unit{M_\odot}$, such that many NSs are given the identical mass of $1.278 \unit{M_\odot}$ in the \textit{delayed} prescription \citep[see][Eq.~19]{Fryer+2012}. This means that many NSs are formed with equal masses and hence we see a mass ratio distribution peaked around unity.

@@

\subsubsection{Eccentricity}
The eccentricity distributions show that detectable BHBHs are most likely to be highly eccentric of the three DCOs. This may seem counter-intuitive since neutron stars receive stronger natal kicks, which cause the orbit to become eccentric. However, these stronger kicks often result in disrupted or too-wide binaries in the more weakly bound NSNSs. In contrast, BHBHs can receive strong kicks that impart high eccentricity without disrupting and thus tend to be more eccentric. This effect is compounded by the fact that we can see BHBHs at lower orbital frequencies, meaning that they have not had as much time to circularise and so still have significant eccentricity by the time of the LISA mission.

@@

\subsubsection{Time since formation}
The progenitors of Galactic DCOs in our sample are formed throughout the history of the Milky Way, with more formed at earlier times (see Fig.~\ref{fig:galaxy_schematic}). In contrast, we find that the \textit{detectable} systems are primarily formed recently, with the majority formed in the past $2 \unit{Gyr}$, in addition to a tail of systems out to earlier times.

@@

\subsubsection{Time until merger}
The final panel of Fig.~\ref{fig:fiducial_pdf_distributions} shows the remaining time until merger for each of the DCO types at the start of LISA mission. The distributions are strikingly similar and peak with merger times of around a Myr.

>>>>

Based on our fiducial simulations, we expect that a 4-year LISA mission will detect about \BHBHFourYear{} BHBHs, \BHNSFourYear{} BHNSs and \NSNSFourYear{} NSNSs (see Table~\ref{tab:detection_rates} for confidence intervals). In this section we discuss the properties of the detectable sources based on our fiducial simulation.   
%We first show the distribution of the sources together with the sensitivity curve in Section~\ref{sec:dcos_on_sc}, before exploring the parameter distributions for detectable sources in Section~\ref{sec:fiducial_distributions} and the measurement uncertainties in Section.~\ref{sec:measurement_uncertainties}.
\subsection{The LISA sensitivity curve and the population of detectable sources}\label{sec:dcos_on_sc}
\begin{figure*}[p]
    \centering
    \includegraphics[width=\textwidth]{2_dcos_on_sc.png}
    \caption{The LISA sensitivity curve is shown together with the density distribution of the characteristic strain and the dominant frequency for all detectable sources in our simulations (top) and separated by type (bottom). Contours show the percentage of the population enclosed. The remaining 2\% of the population is shown as dots with a size that scales with the sampling weight. To guide the interpretation, we show reference lines marking where a circular binary would reside for a given distance (diagonal line) and remaining inspiral time (vertical lines), assuming an average chirp mass $\avg{\mathcal{M}_c}$, orientation and sky location. The curved reference line shows the signal expected for an eccentric binary at 30 kpc. The coloured lines in the bottom panels show a contour that encloses 90\% of the population that is circular. LISA verification binaries are overplotted in the top panel (star symbols).  See also Fig.~\ref{fig:dcos_on_sc_ecc_col} and Sec.~\ref{sec:dcos_on_sc} for a discussion.}
    \label{fig:dcos_on_sc}
\end{figure*}
We show the expected distribution of detectable DCOs together with the LISA sensitivity curve in Fig.~\ref{fig:dcos_on_sc} for all sources combined in the top panel and separated by type in the three lower panels. For reference, we overplot dotted lines to indicate where a binary system would reside for a given distance (diagonal lines) and a remaining inspiral time (vertical lines), assuming they are circular and the average chirp mass of the sample ($\avg{\mathcal{M}_c}$, annotated in each panel), unless indicated otherwise. We observe several features worthwhile discussing, some that earlier studies have touched upon (see Sec.~\ref{sec:compare_studies}), as well as some new findings related to eccentricity.

Firstly, we note that the detectable population is concentrated on the left side of LISA's sensitivity window. The peak is located at a frequency of about $0.4 \unit{mHz}$, which is much lower than the frequency where LISA will be most sensitive (about $10 \unit{mHz}$). This can be understood as an evolutionary effect. DCOs typically form with wide orbits (low frequencies) that would not be detectable yet. Their orbits shrink as they lose angular momentum in the form of gravitational waves leading to an increase of their orbital frequency until they become detectable. Even though the signal emitted by a more compact binary is stronger, such systems are increasingly rare because they evolve faster and faster towards higher frequency as the inspiral accelerates. The vertical grid lines show these rapidly decreasing inspiral times at increasing frequencies. Most of the population is thus expected to reside at low frequencies.

We further note that there is an empty band that separates the LISA sensitivity curve and the detectable population. This is the direct result of the criteria for detection where we require $\rm{SNR} > 7$. For circular systems the SNR is directly given by the height of a system above the curve. For eccentric systems this is also approximately true because, though we have plotted them at the frequency of the dominant harmonic, the y-coordinate indicates the signal from \textit{all} harmonics.

Comparing the lower three panels, one can observe the shift in the frequency at which the peak is located, at $0.3 \unit{mHz}$, $0.7 \unit{mHz}$ and $1 \unit{mHz}$ for BHBH, BHNS and NSNS systems respectively. This is a result of the difference in chirp mass. A higher mass system can emit at lower frequency and still produce the minimum SNR needed for detection. This effect can be used to distinguish the heavier DCOs that we discuss in this work from WDWD systems, at least probabilistically \citep[e.g.][]{Sesana+2020}. In the same way, this also explains the offset in frequency between the population of sources we predict and the LISA verification binaries (star symbols, \citealt{Kupfer+2018}).

Finally we note that the peak of the density distribution of observable sources coincides with the reference line for circular systems at $8 \unit{kpc}$, as expected since this is the distance to the centre of the Milky Way. As can be seen best in the lower panels, the reference lines for $0.1$ and $30 \unit{kpc}$ enclose the majority of systems as expected given the dimensions of our Milky Way. There is, however, a distinct subpopulation of binaries that extend downwards to ASDs of $10^{-19} \unit{Hz}^{-1/2}$, especially around $2 \unit{mHz}$. This offshoot is almost uniquely composed of eccentric binaries, as can be seen most clearly in Fig.~\ref{fig:dcos_on_sc_ecc_col} in our appendix. This can also be seen in Fig.~\ref{fig:dcos_on_sc} from the contour that encompasses 90\% of the \textit{circular} sources in our population, where systems in the offshoot are missing. We also provide a curved reference line that indicates the signal emitted by an eccentric binaries with $e = 0.98$ at a distance of $30 \unit{kpc}$ and average chirp mass. We therefore show that eccentric sources are located in a different region on the sensitivity curve.

\subsection{Properties of the detectable systems}\label{sec:fiducial_distributions}
In Figure~\ref{fig:fiducial_pdf_distributions}, we show kernel density estimators of the 1D distribution of the individual parameters of the population of detectable binaries together with the 1- and 2-$\sigma$ uncertainties (obtained via bootstrapping). We have applied a simple correction for edge effects due to bounded support by mirroring the sample \citep{Schuster+1985}.

\paragraph{Orbital Frequency} The orbital frequency distributions for BHBHs, BHNSs and NSNSs peak at progressively increasing frequencies. As already mentioned above, this is because a higher mass DCO at the same distance and eccentricity requires a lower frequency to produce the same signal-to-noise ratio and thus be detected. The distributions appear nearly symmetric, but closer inspections shows that the left hand side is more populated, which can be seen most clearly in the curve for the BHBHs. This is because of the contribution of highly eccentric binaries, which are most abundant in the BHBH population. These systems are still detectable by LISA as the high eccentricity means that the majority of the GW signal is emitted at higher harmonics, therefore concentrating it at higher frequencies that are located in the LISA band.

@@

\paragraph{Black Hole Mass}
For both the BHBHs and BHNSs, the black hole mass distribution extends to relatively low masses, with $\BHBHmBHBelowEleven{}$ and $\BHNSmBHBelowEleven{}$ respectively below $11 \unit{M_{\odot}}$. Unlike ground-based detectors, LISA is not biased to higher masses and so the mass distribution more closely follows the IMF. In addition, at the high metallicities in the Milky Way, stellar winds are much stronger and strip away much of the stellar mass before BH formation, resulting the less massive black holes. The mass distribution extends down to $2.5 \unit{M_{\odot}}$, our fiducial maximum neutron star mass. Since the \citet{Fryer+2012} \textit{delayed} remnant mass prescription (applied in this work) does not produce a mass gap between neutron stars and black holes. Indeed we expect $\BHBHLowerMassGap{}$ and $\BHNSLowerMassGap{}$ of detected BHBH and BHNS systems to contain a black hole in the lower mass gap. Therefore, LISA could help to confirm or rule out the existence of the lower mass gap.

@@

\paragraph{Mass Ratio} The mass ratio distribution for each DCO type is relatively distinct from the others. The majority of NSNSs have a mass ratio close to unity, with $\NSNSqAbovePointEight{}$ of systems having $q > 0.8$. The reason for the concentration around equal masses is that most NSs are formed either through electron-capture supernovae or from low mass stars. We set the remnant mass for any system formed through ECSN to $1.26 \unit{M_{\odot}}$ (see Sec.~\ref{sec:fiducial_physics}). The remnant mass prescription that we use gives a fixed fallback mass for any star with a CO core mass less than $2.5 \unit{M_\odot}$, such that many NSs are given the identical mass of $1.278 \unit{M_\odot}$ in the \textit{delayed} prescription \citep[see][Eq.~19]{Fryer+2012}. This means that many NSs are formed with equal masses and hence we see a mass ratio distribution peaked around unity.

@@

\paragraph{Eccentricity} The eccentricity distributions show that detectable BHBHs are most likely to be highly eccentric of the three DCOs. This may seem counter-intuitive since neutron stars receive stronger natal kicks, which cause the orbit to become eccentric. However, these stronger kicks often result in disrupted or too-wide binaries in the more weakly bound NSNSs. In contrast, BHBHs can receive strong kicks that impart high eccentricity without disrupting and thus tend to be more eccentric. This effect is compounded by the fact that we can see BHBHs at lower orbital frequencies, meaning that they have not had as much time to circularise and so still have significant eccentricity by the time of the LISA mission.

@@

\paragraph{Time since formation} The progenitors of Galactic DCOs in our sample are formed throughout the history of the Milky Way, with more formed at earlier times (see Fig.~\ref{fig:galaxy_schematic}). In contrast, we find that the \textit{detectable} systems are primarily formed recently, with the majority formed in the past $2 \unit{Gyr}$, in addition to a tail of systems out to earlier times.

@@

\paragraph{Time until merger} The final panel of Fig.~\ref{fig:fiducial_pdf_distributions} shows the remaining time until merger for each of the DCO types at the start of LISA mission. The distributions are strikingly similar and peak with merger times of around a Myr.
